pirouette: Determining the error BEAST2 makes in inferring a phylogeny

### 2019-03-11 article.tex
Can run example 1 on Peregrine
Use new figures
Changes made by @Giappo

\documentclass{article}

@@

\usepackage{listings}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{verbatim}
\usepackage{pgf}
\usepackage{bm}
\usepackage{multirow}
\usepackage{amsfonts}
\usepackage{array}
\usepackage{array}
\usepackage{booktabs}
\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\usepackage{longtable}

@@

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment of characters (usually nucleotides) 
    and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameter 
    estimates. An important ingredient of BEAST2 is 
    the model for the species tree prior, 
    and the tool can in principle be used for inference of 
    macroevolutionary processes underlying the species tree. 
    However, only models that allow for fast computation of 
    the prior probability are possible in practice. 
    This begs the question how accurate the tree estimation is 
    when the real macroevolutionary processes are different 
    from those assumed in BEAST2. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a known/true 
    phylogeny, for example generated by a new 
    macroevolutionary diversification model. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples. \\
\end{abstract}

@@

Because BEAST performs a Bayesian analysis, 
it does not only need character data but also tree priors, 
describing how the diversification occurs, to return posterior phylogenies.
BEAST uses standard tree priors such as Yule or 
constant-rate birth-death [\cite{nee1994reconstructed}] models.
These models are most commonly picked in biological research,
as these are of a desired minimum biological complexity, 
yet computationally fast enough.
The recent development of BEAST2 [\cite{bouckaert2014beast}] provides, 
unlike its predecessor, the possibility for third-party users 
to use their own new or preferred diversification model, 
if an algorithm is provided to calculate the probability 
of a tree given such a diversification model.
There have been plenty of these models (and their associated probability algorithms) 
developed, among others, those in which diversification is 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or that can shift [\cite{etienne2012conceptual}, \cite{rabosky2014automatic}, \cite{alfaro2009nine}, the SLS model. Laudanno et al., in preparation].
Other birth-death models treat speciation as a process that takes time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}], or that happens in bursts of simultaneous events [Laudanno et al., in preparation], or that
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
or even allows concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Note all of these diversification models are available in BEAST2, 
but the option is available.

@@

\verb;pirouette; is very flexible and allows the user 
to specify a big variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in Tables~\ref{tab:definitions} and \ref{tab:options}.

Although many possible tests can be performed, 
we present here only four example research questions (ERQ) that \verb;pirouette; can answer. 
For these ERQs we will estimate the error made by BEAST2 when 
using a birth-death tree prior, in recovering an original tree 
generated under a Yule generative prior.

\begin{sidewaystable}
%\begin{table}
\centering
  %\begin{tabular}{| c | c | l | l |}
  \begin{tabular}{|@{}c|p{5.5cm}|p{2.85cm}@{}|}
    \hline
    \textbf{Symbol} & \textbf{Sub-argument} & \textbf{Possible values} \\ 
    \hline
    $\mathit{tp}$ & 'tree prior', i.e a macroevolutionary speciation model & Yule, BD, CBS, CCP, CEP \\
    $\mathit{cm}$ & 'clock model' for the DNA mutation rates & strict, relaxed log-normal \\
    $\mathit{sm}$ & 'site model' for nucleotide substitution & JC, HKY, TN, GTR \\
    $\mathit{mr}$ & 'mutation rate' & \\
    $\mathit{rs}$ & 'root sequence' of nucleotides at the root of the tree & \\
    $\mathit{mp}$ & 'MRCA prior' knowledge of a most recent common ancestor & \\
    $\mathit{ms}$ & 'MCMC settings' i.e. Markov Chain Monte Carlo algorithm setup & \\
    $\mathit{et}$ & 'experiment type' & Generative, candidate \\
    $\mathit{ri}$ & 'run if' & Always, Best candidate\\
    $\mathit{ef}$ & 'error function' that specifies how to measure the error & nLTT, $|\gamma|$ \\
    $\mathit{tm}$ & 'twin model' & Yule, BD \\
    \hline
  \end{tabular}
  \caption{
    Most important parameter options. BD = birth death, CBS = coalescent
    Bayesian Skyline, CCP = coalescent constant-population, CEP = coalescent
    exponential-population, JC = Jukes-Cantor 1969, HKY = Hasegawa, Kishino and 
    Yano 1985, TN = Tamura 1992, GTR = Generalised time-reversible 
    model (Tavaré 1986)
  }
  \label{tab:options}
%\end{table}
\end{sidewaystable}

\begin{sidewaystable}
%\begin{table}
\centering
  \begin{tabular}{|@{}c|p{6cm}|p{2.85cm}@{}|}
    \hline
    \textbf{Symbol} & \textbf{Macro-argument} & \textbf{Composed by}\\ 
    \hline
    $\mathit{P}$ & 'Generative prior' under which the 'true' tree is simulated & $\mathit{tp}$ \\
    $\mathit{A}$ & 'Alignment model' used for simulation & $\mathit{cm}, \mathit{sm}, \mathit{mr}, \mathit{rs}$ \\
    $\mathit{G}$ & 'Generative model' & $\mathit{P},\mathit{A}$ \\
    $\mathit{I}$ & 'Inference model' to run BEAST2 & $\mathit{cm},\mathit{sm},\mathit{tp},\mathit{mp},\mathit{ms}$ \\
    $\mathit{C}$ & 'Inference conditions' under which $\mathit{I}$ is used in the inference & $\mathit{et}, \mathit{ri}$\\
    $\mathit{X}$ & 'Experiment' & $\mathit{I},\mathit{C}$ \\
    $\mathit{E}$ & 'Error measurement parameters' & $\mathit{ef}$\\
    $\mathit{T}$ & 'Twinning parameters' to set the twin creation procedure & $\mathit{tm}$\\
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in this article.
  }
  \label{tab:definitions}
%\end{table}
\end{sidewaystable}

\iffalse
\begin{table}
\centering
  \begin{tabular}{|@{}c|p{2.8cm}|p{6cm}@{}|}
    \hline
    \textbf{Symbol} & \textbf{Name} & \textbf{Definition} \\ 
    \hline
    $\mathit{t}$ & Tree prior & Macroevolutionary speciation model \\
    $\mathit{c}$ & Clock model & Clock for the DNA mutation rates\\
    $\mathit{s}$ & Site model & Nucleotide substitution model\\
    $\mathit{\mu}$ & Mutation rate & Pace at which mutation occurs, $\mu \in \mathit{s}$ \\
    $\mathit{r}$ & Root sequence & DNA sequence at the root of the tree\\
    $\mathit{m}$ & MRCA prior & Prior knowledge of a most recent common ancestor \\
    $\mathit{v}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup \\
    \hline
    $\mathit{P}$ & Generative prior & Diversification model under which the 'true' tree is simulated \\
    $\mathit{A}$ & Alignment model & Alignment simulation model. It is a combination of $(\mathit{r}, \mathit{\mu}, \mathit{c}, \mathit{s})$, where $\mathit{c}$ = strict \\
    $\mathit{G}$ & Generative model & Combination of a known $\mathit{P}$ and chosen (thus known) $\mathit{A}$ \\
    $\mathit{I}$ & Inference model & Bayesian phylogenetic inference model, a combination of
    $(\mathit{c},\mathit{s},\mathit{t},\mathit{m},\mathit{v})$\\
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ is used in inference \\
    $\mathit{X}$ & Experiment & Combination of $(\mathit{I},\mathit{C})$\\
    $\mathit{E}$ & Error measurement parameters & Errors measurement setup \\
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure \\
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in this article.
  }
  \label{tab:definitions}
  \richel{I've some set notations in some spots, as I am unsure if you like it}
  \giovanni{It would be nice to specify (in the first part of the table) 
    the basic elements composing all the elements of the second part of the table. 
    This could be probably done for "Alignment parameters", "Inference conditions" 
    and "Twinning parameters".
  }
  \richel{
    I feel the next table (which was also your suggestion), does so
    at the right amount of detail
  }
  \giovanni{
    How do we specify what constitutes 'twinning parameters'?
  }
  \richel{Perhaps you'd enjoy my definition?}
\end{table}
\fi

\iffalse
% previous version
\begin{table}
\centering
  \begin{tabular}{|@{}c|p{2.8cm}|@{}l|p{5.5cm}@{}|}
    \hline
    \textbf{Symbol} & \textbf{Name} & \textbf{Definition} \\ 
    \hline
    $\mathit{t}$ & Tree prior & Macroevolutionary speciation model \\
    $\mathit{c}$ & Clock model & Clock for the DNA mutation rates\\
    $\mathit{s}$ & Site model & Nucleotide substitution model\\
    $\mathit{\mu}$ & Mutation rate & Pace at which mutation occurs, $\mu \in \mathit{s}$ \\
    $\mathit{r}$ & Root sequence & DNA sequence at the root of the tree\\
    $\mathit{m}$ & MRCA prior & Prior knowledge of a most recent common ancestor\\
    $\mathit{v}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup \\
    \hline
    $\mathit{P}$ & Generative prior & Diversification model the 'true' tree is simulated with \\
    $\mathit{A}$ & Alignment model & Alignment simulation model, $\mathit{A} \ni \{ \mathit{r}, \mathit{\mu}, \mathit{c}, \mathit{s} \}$, in which $\mathit{c}$ = strict \\
    $\mathit{G}$ & Generative model & Combination of a known $\mathit{P}$ and chosen (thus known) $\mathit{A}$ \\
    $\mathit{I}$ & Inference model & Bayesian phylogenetic inference model, a combination of
    $(\mathit{c},\mathit{s},\mathit{t},\mathit{m},\mathit{v})$\\
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ is used in inference \\
    $\mathit{X}$ & Experiment & Combination of $(\mathit{I},\mathit{C})$\\
    $\mathit{E}$ & Error measurement parameters & Errors measurement setup \\
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure \\
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in this article.
  }
  \label{tab:definitions}
  \richel{I've some set notations in some spots, as I am unsure if you like it}
  \giovanni{It would be nice to specify (in the first part of the table) 
    the basic elements composing all the elements of the second part of the table. 
    This could be probably done for "Alignment parameters", "Inference conditions" 
    and "Twinning parameters".
  }
  \richel{
    I feel the next table (which was also your suggestion), does so
    at the right amount of detail
  }
  \giovanni{
    How do we specify what constitutes 'twinning parameters'?
  }
  \richel{Perhaps you'd enjoy my definition?}
\end{table}
\fi

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a) simulated under the selected generative prior $\mathit{P}$, that is converted to an alignment (2a) according to the alignment model $\mathit{A}$. The user supplies one or more experiments. Every experiment $\mathit{X}$ is defined as a combination of an inference model $\mathit{I}$ and the conditions $\mathit{C}$ under which it is actually used in the inference. Based on the experiments and possibly the alignment (2a), experiments are selected to be actually used in the next step (3a).
    Selected experiments use the alignment (2a) and their inference model $\mathit{I}$ to create a BEAST2 posterior of (parameter estimates and) phylogenies (4a). The posterior trees are compared to the true phylogeny (1a) using the selected error measure $\mathit{E}$, in our case the nLTT statistic [\cite{janzen2015approximate}]. Taking the differences between the true tree's nLTT and each posterior's nLTT results in an error distribution, here displayed as a histogram
    of error values (5a). Optionally, a twin phylogeny (1b) can be generated from the true phylogeny (1a), which then results in an error distribution (5b) with similar intermediate steps.
  }
  \label{fig:pipeline}
  \giovanni{add reference to the generative prior in the figure}
\end{figure}

@@

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
starts from experiments as set up by the user. 
We define an experiment as the combination of an inference model 
and the conditions to actually use it in the inference step.
The inference model can be selected in two ways, either using the
generative model or selecting one or more candidates out of a list of candidate inference models.
An experiment can use the generative model as the inference model, if the generative model is known.
An experiment uses a candidate model if either the generative model is unknown,
or there is an interest in seeing how well non-generative models compete with
the generative model.
Typically, a generative model and/or a best candidate model are used in the actual inference.
When using an experiment with a generative model,
it must use the same site and clock model used in the alignment simulation,
as well as the tree prior that underlied the creation of the given phylogeny 
\giovanni{
  the tree prior appears not to be part of the alignment model. 
  This means that should be part of the generative prior. 
  Is it true? In case we can refer explicitly to it, 
  according to our definitions in table \ref{tab:definitions}
}
\richel{
  Indeed, a tree prior is not part of the alignment model.
  Go ahead and show how you'd like to refer explicitly to the
  table!
}
. 
When using experiments with candidate models,
\verb;pirouette; selects the best candidate based
on the evidence (i.e. marginal likelihood) for the model given the alignment.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac,
therefore the selection of candidate models is not available directly on Windows. 
However, even on Windows systems there is the option to perform the model selection from browser, using \verb;mcbette;, yet the computation for this model selection is
restricted to one hour for free users of the host (Travis CI) continuous
integration service.

@@

An optional step is to generate a 'twin tree' 
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree will be the true tree 
converted to follow the tree prior of the generative model. 
We do so, by finding the parameters 
(e.g. speciation and extinction rate for a birth-death model) 
with the maximum likelihood, 
conditioned on having the same number of tips as the original tree.
We use these parameters to simulate the branching times of the twin tree, 
which we combine with the original tree's topology.

Using a twin tree serves as a control: both the given/true tree
and twin tree will produce an error distribution.
For the true tree, this error distribution is caused by
the shape of the phylogeny (which is different from the tree priors used), 
the stochastic alignment simulation and stochastic Bayesian inference.  
The twin tree is created from the given phylogeny,
as it has the same topology, 
yet its branching times are simulated according to 
the (implemented) generative tree prior.
Due to this, this error distribution for the twin tree is caused 
only by the stochastic alignment simulation and stochastic Bayesian inference.
The difference between these two error distribution is the error caused
by the new/unimplemented tree prior.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Installation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\RequirePackage{rotating}
\documentclass{article}

@@

\usepackage{listings}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{verbatim}
\usepackage{pgf}
\usepackage{bm}
\usepackage{multirow}
\usepackage{amsfonts}
\usepackage{array}
\usepackage{array}
\usepackage{booktabs}
\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\usepackage{longtable}
\usepackage{rotating}
\usepackage{mathtools}

@@

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment of characters (usually nucleotides) 
    and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameter 
    estimates. An important ingredient of BEAST2 is 
    the model for the species tree prior, 
    and the tool can in principle be used for inference of 
    macroevolutionary processes underlying the species tree. 
    However, only models that allow for fast computation of 
    the prior probability are possible in practice. 
    This begs the question of how accurate the tree estimation is 
    when the real macroevolutionary processes are different 
    from those assumed in BEAST2. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a known/true 
    phylogeny, for example generated by a new 
    macroevolutionary diversification model. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples. \\
\end{abstract}

@@

Because BEAST performs a Bayesian analysis, 
it does not only need character data but also tree priors, 
describing how the diversification occurs, to return posterior phylogenies.
BEAST uses standard tree priors such as Yule or 
constant-rate birth-death [\cite{nee1994reconstructed}] models.
These models are most commonly picked in biological research,
as these are of a desired minimum biological complexity, 
yet computationally fast enough.
The recent development of BEAST2 [\cite{bouckaert2014beast}] provides, 
unlike its predecessor, the possibility for third-party users 
to use their own new or preferred diversification model, 
if an algorithm is provided to calculate the probability 
of a tree given such a diversification model.
There have been plenty of these models (and their associated probability algorithms) 
developed, among others, those in which diversification is 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or that can shift [\cite{etienne2012conceptual}, \cite{rabosky2014automatic}, \cite{alfaro2009nine}, the SLS model. Laudanno et al., in preparation].
Other birth-death models treat speciation as a process that takes time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}], or that happens in bursts of simultaneous events [Laudanno et al., in preparation], or that
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
or even allows concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Not all of these diversification models are available in BEAST2, 
but the option is available.

@@

\verb;pirouette; is very flexible and allows the user 
to specify a big variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in Tables~\ref{tab:options} and \ref{tab:definitions}.

Although many possible tests can be performed, 
we present here only four \giovanni{five?} example research questions (ERQ) that \verb;pirouette; can answer. 
For these ERQs we will estimate the error made by BEAST2 when 
using a birth-death tree prior, in recovering an original tree 
generated under a Yule generative prior.

\begin{sidewaystable}
\centering
  \begin{tabular}{|@{}c|p{2.5cm}|p{9cm}|p{4.5cm}@{}|}
    \hline
    \centering
    \textbf{Symbol} & \textbf{Sub-argument} & \textbf{Description} & \textbf{Possible values} \\ 
    \hline
    $\mathit{t_{p}}$ & tree prior & Macroevolutionary speciation model & Yule, BD, CBS, CCP, CEP \\
    $\mathit{c_{m}}$ & clock model & Clock for the DNA mutation rates & strict, relaxed log-normal \\
    $\mathit{s_{m}}$ & site model & Nucleotide substitution model & JC, HKY, TN93, GTR \\
    $\mathit{m_{r}}$ & mutation rate & Pace at which mutation occurs & $m_{r} \in \mathbb{R}_{>0}$\\
    $\mathit{r_{s}}$ & root sequence & DNA sequence at the root of the tree & any combination of a, c, g, t \\
    $\mathit{m_{p}}$ & MRCA prior & Prior knowledge of a most recent common ancestor & see documentation \\
    $\mathit{m_{s}}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup & see documentation \\
    $\mathit{m_{t}}$ & model type & Criterion to select an inference model & Generative, Candidate \\
    $\mathit{r_{i}}$ & run if & Condition under which an inference model is used & Always, Best candidate\\
    $\mathit{e_{f}}$ & error function & Specifies how to measure the error & nLTT, $|\gamma|$ \\
    $\mathit{t_{m}}$ & twin model & Sets the tree prior for the twin tree & Yule, BD \\
    \hline
  \end{tabular}
  \caption{
    Most important parameter options. BD = birth death, CBS = coalescent
    Bayesian Skyline, CCP = coalescent constant-population, CEP = coalescent
    exponential-population, JC = Jukes-Cantor 1969 (\cite{jukes1969evolution}), HKY = Hasegawa, Kishino and 
    Yano (\cite{hasegawa1985dating}), TN93 = \cite{tamura1993estimation}, GTR = Generalised time-reversible 
    model (\cite{tavare1986some}).
  }
  \label{tab:options}
\bigskip

  \begin{tabular}{|@{}c|p{4cm}|p{9cm}|p{3cm}@{}|}
    \hline
    \centering
    \textbf{Symbol} & \textbf{Macro-argument} & \textbf{Description} & \textbf{Includes} \\
    \hline
    $\mathit{G}$ & Generative prior & Diversification model to generate the 'true' tree & $\mathit{t_{p}}$ \\
    $\mathit{A}$ & Alignment model & Model used to simulate the alignment ($\mathit{c_{m}}$ must be strict) & $\mathit{c_{m}}, \mathit{s_{m}}, \mathit{m_{r}}, \mathit{r_{s}}$ \\
    $\mathit{I}$ & Inference model & Phylogenetic inference model to run BEAST2 & $\mathit{c_{m}},\mathit{s_{m}},\mathit{t_{p}},\mathit{m_{p}},\mathit{m_{s}}$ \\
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ is used in the inference & $\mathit{m_{t}}, \mathit{r_{i}}$\\
    $\mathit{X}$ & Experiment & Full setting for the BEAST2 run & $\mathit{I},\mathit{C}$ \\
    $\mathit{E}$ & Error measure parameters & Errors measurement setup & $\mathit{e_{f}}$\\
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure & $\mathit{t_{m}}$\\
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in the main text and in Fig~\ref{fig:pipeline}.
  }
  \label{tab:definitions}
  \giovanni{Should we rename "Twinning Parameters" to "Twinning model" and "twin model" to "twin prior"?}
\end{sidewaystable}

\iffalse
\begin{table}
\centering
  \begin{tabular}{|@{}c|p{2.8cm}|p{6cm}@{}|}
    \hline
    \textbf{Symbol} & \textbf{Name} & \textbf{Definition} \\ 
    \hline
    $\mathit{t}$ & Tree prior & Macroevolutionary speciation model \\
    $\mathit{c}$ & Clock model & Clock for the DNA mutation rates\\
    $\mathit{s}$ & Site model & Nucleotide substitution model\\
    $\mathit{\mu}$ & Mutation rate & Pace at which mutation occurs, $\mu \in \mathit{s}$ \\
    $\mathit{r}$ & Root sequence & DNA sequence at the root of the tree\\
    $\mathit{m}$ & MRCA prior & Prior knowledge of a most recent common ancestor \\
    $\mathit{v}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup \\
    \hline
    $\mathit{P}$ & Generative prior & Diversification model under which the 'true' tree is simulated \\
    $\mathit{A}$ & Alignment model & Alignment simulation model. It is a combination of $(\mathit{r}, \mathit{\mu}, \mathit{c}, \mathit{s})$, where $\mathit{c}$ = strict \\
    $\mathit{G}$ & Generative model & Combination of a known $\mathit{P}$ and chosen (thus known) $\mathit{A}$ \\
    $\mathit{I}$ & Inference model & Bayesian phylogenetic inference model, a combination of
    $(\mathit{c},\mathit{s},\mathit{t},\mathit{m},\mathit{v})$\\
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ is used in the inference \\
    $\mathit{X}$ & Experiment & Combination of $(\mathit{I},\mathit{C})$\\
    $\mathit{E}$ & Error measurement parameters & Errors measurement setup \\
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure \\
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in this article.
  }
  \label{tab:definitions}
  \richel{I've some set notations in some spots, as I am unsure if you like it}
  \giovanni{It would be nice to specify (in the first part of the table) 
    the basic elements composing all the elements of the second part of the table. 
    This could be probably done for "Alignment parameters", "Inference conditions" 
    and "Twinning parameters".
  }
  \richel{
    I feel the next table (which was also your suggestion), does so
    at the right amount of detail
  }
  \giovanni{
    How do we specify what constitutes 'twinning parameters'?
  }
  \richel{Perhaps you'd enjoy my definition?}
\end{table}
\fi

\iffalse
% previous version
\begin{table}
\centering
  \begin{tabular}{|@{}c|p{2.8cm}|@{}l|p{5.5cm}@{}|}
    \hline
    \textbf{Symbol} & \textbf{Name} & \textbf{Definition} \\ 
    \hline
    $\mathit{t}$ & Tree prior & Macroevolutionary speciation model \\
    $\mathit{c}$ & Clock model & Clock for the DNA mutation rates\\
    $\mathit{s}$ & Site model & Nucleotide substitution model\\
    $\mathit{\mu}$ & Mutation rate & Pace at which mutation occurs, $\mu \in \mathit{s}$ \\
    $\mathit{r}$ & Root sequence & DNA sequence at the root of the tree\\
    $\mathit{m}$ & MRCA prior & Prior knowledge of a most recent common ancestor\\
    $\mathit{v}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup \\
    \hline
    $\mathit{G}$ & Generative prior & Diversification model the 'true' tree is simulated with \\
    $\mathit{A}$ & Alignment model & Alignment simulation model, $\mathit{A} \ni \{ \mathit{r}, \mathit{\mu}, \mathit{c}, \mathit{s} \}$, in which $\mathit{c}$ = strict \\
    $\mathit{G}$ & Generative model & Combination of a known $\mathit{P}$ and chosen (thus known) $\mathit{A}$ \\
    $\mathit{I}$ & Inference model & Bayesian phylogenetic inference model, a combination of
    $(\mathit{c},\mathit{s},\mathit{t},\mathit{m},\mathit{v})$\\
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ is used in inference \\
    $\mathit{X}$ & Experiment & Combination of $(\mathit{I},\mathit{C})$\\
    $\mathit{E}$ & Error measurement parameters & Errors measurement setup \\
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure \\
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in this article.
  }
  \label{tab:definitions}
  \richel{I've some set notations in some spots, as I am unsure if you like it}
  \giovanni{It would be nice to specify (in the first part of the table) 
    the basic elements composing all the elements of the second part of the table. 
    This could be probably done for "Alignment parameters", "Inference conditions" 
    and "Twinning parameters".
  }
  \richel{
    I feel the next table (which was also your suggestion), does so
    at the right amount of detail
  }
  \giovanni{
    How do we specify what constitutes 'twinning parameters'?
  }
  \richel{Perhaps you'd enjoy my definition?}
\end{table}
\fi

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a) simulated under the selected generative prior $\mathit{P}$, that is converted to an alignment (2a) according to the alignment model $\mathit{A}$. The user supplies one or more experiments. Every experiment $\mathit{X}$ is defined as a combination of an inference model $\mathit{I}$ and the conditions $\mathit{C}$ under which it is actually used in the inference. Based on the experiments and possibly the alignment (2a), experiments are selected to be actually used in the next step (3a).
    Selected experiments use the alignment (2a) and their inference model $\mathit{I}$ to create a BEAST2 posterior of (parameter estimates and) phylogenies (4a). The posterior trees are compared to the true phylogeny (1a) using the selected error measure $\mathit{E}$, in our case the nLTT statistic [\cite{janzen2015approximate}]. Taking the differences between the true tree's nLTT and each posterior's nLTT results in an error distribution, here displayed as a histogram
    of error values (5a). Optionally, a twin phylogeny (1b) can be generated from the true phylogeny (1a), which then results in an error distribution (5b) with similar intermediate steps.
  }
  \label{fig:pipeline}
  \giovanni{add reference to the generative prior P (or G) in the figure}
  \giovanni{I think that a list of possible models should replace the "experiment" cake in the middle, then C should be drawn on the arrow connecting the figure in the middle to 3a. "I" should replace the circled x of passage 3a. An additional C can be drawn alongside the arrow 2a -> 3a. After that a rectangle can be drawn around the whole middle section (including 3a, 3b and the middle C) to denote that this is phase "X".}
  \giovanni{We might also need an arrow going from the central figure to the top arrow (the one with "T") to specify that the twin is generated according to the chosen inference model.}
\end{figure}

@@

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
starts from experiments as set up by the user. 
We define an experiment as the combination of an inference model 
and the conditions to actually use it in the inference step.
The inference model can be selected in two ways, either using the
generative model or selecting one or more candidates out of a list of candidate inference models.
An experiment can use the generative model as the inference model, if the generative model is known.
An experiment uses a candidate model if either the generative model is unknown,
or there is an interest in seeing how well non-generative models compete with
the generative model.
Typically, a generative model and/or a best candidate model are used in the actual inference.
When using an experiment with a generative model,
it must use the same site and clock model used in the alignment simulation,
as well as the tree prior that underlied the creation of the given phylogeny. 
When using experiments with candidate models,
\verb;pirouette; selects the best candidate based
on the evidence (i.e. marginal likelihood) for the model given the alignment.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac,
therefore the selection of candidate models is not available directly on Windows. 
However, even on Windows systems there is the option to perform the model selection from browser, using \verb;mcbette;, yet the computation for this model selection is
restricted to one hour for free users of the host (Travis CI) continuous
integration service.

@@

\iffalse
%version 1
An optional step is to generate a 'twin tree' 
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree will be the true tree 
converted to follow the tree prior of the chosen inference model.

To do so we exploit the likelihood function associated with the chosen inference model. We find the parameters (e.g. speciation and extinction rate for a birth-death model) that maximize such likelihood applied to the true tree, conditioned on its number of tips.

We use these parameters to simulate the branching times of the twin tree, 
which we combine with the original tree's topology.

Using a twin tree serves as a control: both the given/true tree
and twin tree will produce an error distribution.
For the true tree, this error distribution is caused by
the shape of the phylogeny (which is different from the tree priors used), 
the stochastic alignment simulation and stochastic Bayesian inference.  
The twin tree is created from the given phylogeny,
as it has the same topology, 
yet its branching times are simulated according to 
the (implemented) generative tree prior.
Due to this, this error distribution for the twin tree is caused 
only by the stochastic alignment simulation and stochastic Bayesian inference.
The difference between these two error distribution is the error caused
by the new/unimplemented tree prior.
\fi

\iffalse
%version 2 with no formulae
An optional step is to generate a 'twin tree' 
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree. We want to produce such tree 'converting' the original tree, created according to the generative prior $\mathit{G}$, into the chosen inference model $\mathit{I}$.

To do so we exploit the likelihood function associated with the chosen inference model $\mathit{I}$. We find the parameters (e.g. speciation and extinction rate for a birth-death model) that maximize such likelihood applied to the true tree, conditioned on its number of tips.

We use these parameters to simulate an equal amount of branching times for the twin tree simulated under the chosen inference model's process. We then combine these branching times with the original tree's topology.

Using a twin tree serves as a control: both the given/true tree and twin tree will produce an error distribution. We assume that these errors are a natural product of the input tree and the intermediate steps of the pipeline. All the factors that can cause such errors are: the branching times of the tree (as a consequence of the tree prior), the topology of the tree, the stochasticity induced by simulating the alignment (step 1a $\rightarrow$ 2a in Fig.~\ref{fig:pipeline}) and the stochasticity induced by the Bayesian inference (step 2a $\rightarrow$ 4a in Fig.~\ref{fig:pipeline}). By imposing the same topology and replicating the pipeline a sufficient number of times we can sort out all the possible causes that are not a direct consequence of the choice of the tree prior.

Potential differences in shapes of error distributions will therefore be consequences of different tree priors for the two parallel pipelines: the use of prior $\mathit{G}$ for the original pipeline and of prior $\mathit{I}$ for the twin pipeline. In the second case, though, such error distribution is not influenced by the mismatch between how the tree is generated and how it is analyzed by BEAST2.
\fi

\giovanni{This is a slightly more formal version. I personally find it clearer. If you don't like it we can use version 2 (see above in the source). In case you really like it, instead, I can try to make the rest a bit more similar to this.}

%version 3, more formal
An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree, following a parallel pipeline. We want to produce such tree 'converting' the original tree $\tau_{G}$, created according to the generative prior $\mathit{G}$, to be in accordance with the chosen inference model $\mathit{I}$:
\begin{align}
    \tau_{G} = (\Vec{t}_{G}, \psi_{G}) \xrightarrow[]{\mathit{T}} \tau_{I} = (\Vec{t}_{I}, \psi_{I}),
\end{align}
where $\Vec{t}$ are the tree branching times and $\psi$ is tree topology, which provides information on where new lineages branch out at each branching time.
